Parsing Postfix log files

### 2008-01-28 doc/logparser.tex
New command \sectionref{} which puts section squiggles in front of references, and all appropriate references changed to use it.

% Replacement for \ref{}, adds the page number too.
\newcommand{\refwithpage}[1]{%
    \empty{}\ref{#1} [page~\pageref{#1}]%
}

@@

XXX USE \textsection{} INSTEAD OF SECTION WHEN REFERRING TO \\OTHER
SECTIONS\@?

@@

This paper and the program it describes are part of a larger project to
optimise a server's Postfix restrictions, generate statistics and graphs,
and provide a platform on which new restrictions can be trialled and
evaluated to see if they are beneficial in the fight against spam.  The
program parses Postfix logs and populates a database with the data gleaned
from those logs, providing a consistent and simple view of the logs which
future tools can utilise.  The gathered data can then be used to optimise
current anti-spam measures, provide a baseline to test new anti-spam
measures against, or to produce statistics showing how effective those
measures are.\footnote{Section~\refwithpage{introduction} lists some more
example uses.}

@@

Postfix is a \MTA{} with the following design aims (in order of
importance): security, flexibility of configuration, scalability, and high
performance.  It features extensive, extensible, optional anti-spam
restrictions, allowing an administrator to deploy those restrictions which
they judge suitable for their site's needs, rather than a fixed set chosen
by Postfix's author.  These restrictions can be selectively applied,
combined and bypassed on a per-client, per-recipient or per-sender basis,
allowing varying levels of severity and/or permissiveness.  Postfix
leverages simple lookup tables to support arbitrarily complicated
user-defined sequences of restrictions and exceptions, with policy
servers\footnote{Policy servers will be explained in
section~\refwithpage{policy servers}.} as the ultimate in flexibility.
Administrators can also supply their own rejection messages to make it
clear to senders why exactly their mail was rejected.  Unfortunately this
flexibility has a cost: complexity in the logs generated.  While it is easy
to use standard Unix text processing utilities to determine the fate of an
individual email, following the journey an email takes through Postfix can
be quite difficult.  The logs tend to follow a 90\%-10\% pattern: 90\% of
the time the journey is simple, but the other 10\% of the time requires
90\% of the code.\footnote{These numbers don't have a solid scientific
basis, they're based on gut feeling from writing and debugging the
software.}

Postfix's design follows the Unix philosophy of \textit{Write programs that
do one thing and do it well\/}~\cite{unix-philosophy}, and is separated
into various component programs to perform the tasks required of an
\MTA{}\@: receive mail, send mail, local delivery of mail, etc. --- full
details can be found at~\cite{postfix-overview}.  Each log line contains
the name of the Postfix program which produced it, and this information is
used when determining which rules should be used to parse each log line
(see section~\refwithpage{rule characteristics} for details.)

@@

It should be clear from the earlier Postfix background
(section~\refwithpage{postfix background}) that logs produced by Postfix
are not fixed; they vary widely from host to host, depending on the set of
restrictions chosen by the administrator.  With this in mind, one of the
parser's design aims was to make adding new rules as easy as possible, to
enable administrators to properly parse their log lines.  The main design
decision which enables this is the separation of rules from the parsing
algorithm, and secondly the division of the algorithm into discrete
actions.

@@

Separating the rules from the algorithm makes it possible to parse new log
lines without modifying the core parsing algorithm.  Although this may seem
like a trivial point, is it substantially more difficult to understand a
program's entire parsing algorithm, identify the correct location to
change, and make the appropriate changes without adversely affecting
parsing, particularly as there may be edge cases which are not immediately
obvious.\footnote{See
section~\refwithpage{yet-more-aborted-delivery-attempts} for a complication
which occurs only four times in 93 log files tested.}  Requiring changes to
the parsing algorithm also complicates upgrades, as the changes must be
preserved during the upgrade, and may clash with changes made by the
developer.  \parsername{} allows the user to add new rules to the database
without changing the parsing algorithm, unless the new log lines to be
parsed require functionality not already provided by the algorithm.  If the
new log lines do require new functionality, new actions can be added to the
parser without modifying existing actions or other parts of the algorithm;
only in the rare case that the new actions require support from other
sections of the code will more extensive changes be required.

@@

Parsing in \LMA{} requires major changes to the code to parse new log lines
or extract additional data.  It does not appear to deal with any of the
complications discussed in sections~\refwithpage{complications} (initial
complications) and~\refwithpage{additional complications} (additional
complications).  It does not differentiate between different types of
rejections, so it is not suitable for the purposes of this project; the
data about which restriction caused the rejection is discarded, whereas an
important goal of this project is to retain that data to aid optimisation
and evaluation of Postfix restrictions.  \LMA{} fails to parse Postfix log
files generated on Solaris hosts because the fields added automatically
differ from those added on Linux hosts; logs from Solaris hosts (and
possibly other operating systems) thus require preprocessing before parsing
by \LMA{}.

@@

The database schema can be conceptually divided in two: the rules which are
used to parse log files, and the data saved from the parsing of log files.
Rules have the fields required to parse the log lines, extract data to be
saved, and the action to be executed; they also have several fields which
aid the user in understanding what each rule parses.  The rules are
described in detail in section~\refwithpage{rules} but the fields are
covered here.

The data saved from parsing the logs is also divided into two tables as
described below: connections and results.  The connections table contains a
row for every mail accepted and every connection where there was a
rejection; the individual fields will be described in the forthcoming
section~\refwithpage{connections table}.  The results table has one or more
rows per row in the connections table, depending on the specifics of the
mail/connection; the fields will be covered in detail in
section~\refwithpage{results table}.

@@

Rules are discussed in detain in section~\refwithpage{rules}, but the rules
table is covered here.  Rules are created by the user, not the parser, and
will not be modified by the parser (except for the hits and hits\_total
fields).  Rules parse the individual lines, telling the parser which fields
to extract and what action to take for that line.

@@

        For efficiency the keywords are expanded and every rule's \regex{}
        is compiled before attempting to parse the log file --- otherwise
        each \regex{} would be recompiled each time it was used, resulting
        in a large, data dependent slowdown.  Rule efficiency concerns are
        discussed in section~\refwithpage{rule efficiency}.

@@

    \item [action] The action the algorithm will take; a full list can be
        found in Section~\refwithpage{actions-in-detail}.

@@

    \item [hits] is an efficiency measure.  This counter is maintained for
        every rule and incremented each time the rule successfully matches.
        At the start of each run the program sorts the rules in descending
        order of hits, and at the end of the run updates every rule's hits.
        Assuming that the distribution of log lines is reasonably
        consistent between log files, rules matching more commonly
        occurring log lines will be tried before rules matching less
        commonly occurring log lines, lowering the program's execution
        time.  Rule ordering for efficiency is discussed in
        section~\refwithpage{rule ordering for efficiency}.

@@

    \item [priority] This is the user-configurable companion to hits: rules
        will be tried in order of priority, overriding hits.  This allows
        more specific rules to take precedence over more general rules
        (described in section~\refwithpage{overlapping rules}).

@@

This section discusses the rules used in parsing Postfix log files,
starting with rule characteristics, continuing with detecting and dealing
with overlapping rules, and the problems they can cause.  An example rule
and a line it would match are provided, plus a description of how the
fields in the rule are used when matching and performing the action (the
details of the table containing the rules have already been described in
section~\refwithpage{rule attributes},).  This section continues with a
discussion of rule efficiency concerns, referring to the graphs in
section~\refwithpage{graphs}, and finishes with an explanation of the
algorithm used to generate a new \regex{} from unparsed lines.

Please refer to section~\refwithpage{parser design} for a discussion of why
the rules and actions have been separated in the parser's design.

@@

    \item Compare the results of parsing using sorted, shuffled and
        reversed rules.\footnote{See section~\refwithpage{rule efficiency}
        for more details of sorting the rules.}  Parse a number of log
        files using normal sorting, dump a textual representation of the
        connections and results tables, and delete everything from those
        tables.  Repeat with shuffled and reversed sorting.  If there are
        no overlapping rules the tables from each run will be identical;
        differences indicate overlapping rules.  Which rules overlap can be
        determined by examining the differences in the tables: each result
        contains the id of the rule which created it, so the rule
        referenced in the normal table overlaps with the rule referenced in
        the reversed table.  Unfortunately this method cannot prove the
        absence of overlapping rules; it can detect overlapping rules, but
        only if there are log lines in the input files which match more
        than one rule.

@@

    \item [action:] will be executed if the \regex{} matches successfully
        (see section~\refwithpage{actions-in-detail} for full details).

@@

    \item [hits, hits\_total and priority:] hits and priority are used in
        ordering the rules (see section~\refwithpage{rule ordering for
        efficiency}); hits is set to the number of successful matches at
        the end of the parsing run, and hits\_total has hits added to it,
        but is otherwise unused by the algorithm.

@@

Rule ordering was mentioned in section~\refwithpage{rule attributes} and
will be covered in greater detail in this section.  At the time of writing
there are 139 different rules, with the top 10\% matching the vast majority
of the log lines, and the remaining log lines split across the other 90\%
of the rules (as shown in graph~\refwithpage{rule hits graph}).  Assuming
that the distribution of log lines is reasonably steady over time, program
efficiency should benefit from trying more frequently matching rules before
those which match less frequently.  To test this hypothesis three full test
runs were performed with different rule orderings:

@@

    \item [Pre-process the file]  The new algorithm leverages the knowledge
        gained while writing rules and performs a large number of
        substitutions on the input lines, replacing commonly occurring
        variable terms (e.g.\ email addresses, \IP{} addresses, the
        standard start of rejection messages, etc.) with \regex{} keywords
        which the parser will expand when it loads the rule (see the
        \regex{} entry in section~\refwithpage{rule attributes}).  The
        purpose of this phase is to utilise the existing knowledge to
        create more accurate \regexes{}.  The new log lines are written to
        a temporary file, which all subsequent stages use instead of the
        original input file.

@@

    \item The database table containing the rules is dealt with in
        section~\refwithpage{rule attributes}, and is not duplicated in
        this section.

@@

Please refer to section~\refwithpage{parser design} for a discussion of why
the rules and actions have been separated in the parser's design.

@@

Multiple independent mails may be delivered during one connection: this
requires cloning the current data as soon as a mail is accepted, so that
subsequent mails won't trample over each other's data.  This must be done
every time a mail is accepted, as it's impossible to tell in advance which
connections will accept multiple mails.  It is quite easy to overlook this
complication because only a small minority of connections accept more than
one mail. Happily once the mail has been accepted log entries won't be
correlated by \pid{} for that mail any more (its queueid will be used
instead), so there isn't any ambiguity about which mail a given log line
belongs to.\footnote{Unfortunately this statement is not completely
accurate: see section~\refwithpage{timeouts-during-data-phase} for details.
However in general there isn't any ambiguity about which data structure
should be used for a given log line.}  The original connection will be
discarded unsaved when the client disconnects if it doesn't have any data
worth saving, i.e.\ no rejections.  One unsolved difficulty is
distinguishing between different groups of rejections, e.g.\ when dealing
with the following sequence:

@@

Figures~\refwithpage{flow chart image part 1} and~\refwithpage{flow chart
image part 2} shows the paths the data representing a mail/connection can
take through the parser algorithm.  The flow chart covers the most common
paths only; there are additional, uncommon paths which are excluded for the
sake of clarity, but are described in section~\refwithpage{additional
complications}.

@@

    Several
more states and corresponding branches are omitted for the sake of clarity
(these extra states are caused by solving the difficulties described in
section~\refwithpage{additional complications}).

The solution to the second last complication encountered
(section~\refwithpage{out of order log lines}) requires that a list of the
acceptable combinations of Postfix programs a mail can pass through;
however this does not correspond to the \textit{actions\/} a mail must pass
through.  Before that complication was overcome, i.e.\ for the majority of
the development of this parser, there was nothing in the parser explicitly
encoding the states or paths a mail could take; the vast majority of the
parser has no need for this flow of data to be specified.  This eases the
process of adding new actions (described in section~\refwithpage{adding new
actions}), as there is no requirement to explicitly insert the action into
an algorithmic version of the flow chart below.

@@

Section~\refwithpage{actions-in-detail} explains the actions in substantive
detail; this section omits such detail because it would clutter and confuse
the algorithm description.  The flow chart in
section~\refwithpage{flow-chart} should also be consulted while reading
this section.  For the sake of clarity this description covers only the
most common paths through the algorithm, as including every path would
hinder understanding; the more uncommon paths are caused by the
complications described in section~\refwithpage{additional complications},
and are covered therein.

@@

    \item If one or more mails were accepted there will be more log entries
        for those mails later, see section~\refwithpage{mail-delivery}.

@@

Re-injection due to forwarding sadly lacks explicit log lines of its
own;\footnote{Previously discussed in section~\refwithpage{complications},
complication 3.} re-injection is somewhat awkward to explain because it
overlaps both the mail acceptance and mail delivery sections, so discussion
is deferred to section~\refwithpage{tracking re-injected mail}.

Internally generated mails lack any explicit origin in Postfix 2.2.x and
must be detected using heuristics (see
section~\refwithpage{identifying-bounce-notifications} for details).
Bounce notifications are the primary example of internally generated mails,
though there may be other types.

@@

    \item Sometimes mail is re-injected and the child mail needs to be
        tracked by the parent mail (action: track; flowchart:~10) ---
        section~\refwithpage{tracking re-injected mail} discusses this in
        detail.

    \item Eventually the mail will be delivered, bounced, or deleted by the
        administrator (action: commit; flowchart:~12).  This is the last
        log line for this particular mail (though it may be indirectly
        referred to if it was re-injected).  If it is neither parent nor
        child of re-injection the data is cleaned up and entered in the
        database (flowchart:~14), then deleted from the state tables.
        Re-injected mails are described in section~\refwithpage{tracking
        re-injected mail}.

@@

It should be reiterated that the actions above happen whether the mail is
delivered to a mailbox, piped to a command, delivered to a remote server,
bounced (due to a mail loop, delivery failure, or five day timeout), or
deleted by the administrator.  The exception is what happens after delivery
to the parent or children of mail re-injected due to forwarding, as
explained in section~\refwithpage{tracking re-injected mail}.

@@

    \item The indicator which more commonly introduces re-injection is when
        \daemon{qmgr} selects a mail with a previously unseen queueid for
        delivery (action: mail\_picked\_for\_delivery; flowchart:~3), in
        which case a new data structure will be created.  The mail will be
        flagged as having unknown origins; this flag should be subsequently
        cleared once the origin has been established.  This may also be an
        indicator that the mail is a bounce notification, see
        section~\refwithpage{identifying-bounce-notifications} for details.

@@

    \item [EXPIRY] If Postfix has not managed to deliver a mail after
        trying for five days it will give up and return the mail to the
        sender.  When this happens the mail will not have a combination of
        Postfix programs which passes the valid combinations check (see
        section~\refwithpage{out of order log lines}).  To ensure that the
        mail can be committed the EXPIRY action sets a flag marking the
        mail as expired; the flag later causes the valid combinations check
        to be skipped, so the mail will be committed.

@@

    \item [MAIL\_PICKED\_FOR\_DELIVERY] This action represents Postfix
        picking a mail from the queue to deliver. This action is used for
        both \daemon{qmgr} and \daemon{cleanup} as it needs to deal with
        out of order log lines; see section~\refwithpage{discarding cleanup
        lines} for details.

@@

    \item [PICKUP] The PICKUP action corresponds to the \daemon{pickup}
        service dealing with a locally submitted mail.  Out of order log
        entries may have caused the state table entry to already exist (see
        section~\refwithpage{pickup logging after cleanup}); otherwise it
        is created.  The data from the log line is then saved to the state
        table entry.

@@

    \item [TIMEOUT] The connection timed out so the mail currently being
        transferred must be discarded. The mail may have been accepted, in
        which case there's a data structure to dispose of, or it may not in
        which case there is not.  See
        section~\refwithpage{timeouts-during-data-phase} for the gory
        details.

@@

Adding new actions is almost as simple as adding new rules, though actions
are not stored in the database.  The implementor writes a subroutine which
accepts the standard arguments given to actions, and registers it as an
action by calling the parser subroutine add\_actions() with the name of the
new action subroutine as a parameter.  No other work is required on the
implementor's part; all of their attention and effort can be focused on the
correctness of their action.  The only negative aspect is that the process
involves editing the parser source code, which makes upgrading to a later
version of the parser more difficult, though by no means impossible.  If
the author of the new action wishes they can take advantage of the parser's
object oriented implementation~\cite{Wikipedia-object-orientation} by
subclassing it and implementing their changes in the derived class,
allowing future upgrading of the parser with greatly reduced chance of
conflicts.\footnote{The real difference between the two approaches is where
the new code is placed.  The simpler option is to change the parser code
directly, but those changes will then have to be made to subsequent
versions of the parser, and as the scope of the changes increases so does
the chance of conflict.  The more time consuming option is to write a
subclass containing the new actions and change the program which invokes
the parser so that it uses the subclass rather than the parser; the changes
required to the program invoking the parser are minor and much less likely
to lead to conflicts when upgrading to future versions of the parser.  An
alternative is to submit new actions to the author of the parser for
inclusion in future versions, resulting in two benefits: the new actions do
not need to be maintained separately, and other users of the parser will
avail of the new functionality.} The action may need to extend the list of
valid combinations described in section~\refwithpage{out of order log
lines} if the addition creates a different set of acceptable programs, but
this is unlikely to occur, as it would require parsing log lines from
Postfix components the parser currently ignores.\footnote{The mail server
used for development does not utilise the \daemon{lmtp} or \daemon{virtual}
delivery agents, so this parser does not have rules to handle log lines
from those components.  Adding new rules to parse their log lines is a
simple process, though if their behaviour differs significantly from the
\daemon{smtp} or \daemon{local} delivery agents new actions may be
required.  The mail server in question is a production mail server handling
mail for a university department; the benefit is that the logs used exhibit
the idiosyncrasies and foibles a mail server in the wild must deal with,
but the downside is that significantly altering the configuration just to
log messages from a different Postfix component is not an option.}

@@

Postfix 2.2.x (and presumably previous versions) lacks explicit logging
when bounce notifications are generated; suddenly there will be log entries
for a mail which lacks an obvious source.  There are similarities to the
problem of re-injected mails discussed in section~\refwithpage{tracking
re-injected mail}, but unlike the solution described therein bounce
notifications do not eventually have a log line which identifies their
source.  Heuristics must be used to identify bounce notifications, and
those heuristics are:

@@

This complication is further complicated by the presence of out of order
\daemon{cleanup} lines: see section~\refwithpage{discarding cleanup lines}
for details.

@@

    \item The appropriate action to take --- discussed in
        section~\refwithpage{mails-covered}.

>>>>

% Replacement for \ref{}, adds the page number too.
\newcommand{\refwithpage}[1]{%
    \empty{}\ref{#1} [page~\pageref{#1}]%
}
% section references, automatically add \textsection
\newcommand{\sectionref}[1]{%
    \textsection{}\refwithpage{#1}%
}

@@

@@

This paper and the program it describes are part of a larger project to
optimise a server's Postfix restrictions, generate statistics and graphs,
and provide a platform on which new restrictions can be trialled and
evaluated to see if they are beneficial in the fight against spam.  The
program parses Postfix logs and populates a database with the data gleaned
from those logs, providing a consistent and simple view of the logs which
future tools can utilise.  The gathered data can then be used to optimise
current anti-spam measures, provide a baseline to test new anti-spam
measures against, or to produce statistics showing how effective those
measures are.\footnote{\sectionref{introduction} lists some more example
uses.}

@@

Postfix is a \MTA{} with the following design aims (in order of
importance): security, flexibility of configuration, scalability, and high
performance.  It features extensive, extensible, optional anti-spam
restrictions, allowing an administrator to deploy those restrictions which
they judge suitable for their site's needs, rather than a fixed set chosen
by Postfix's author.  These restrictions can be selectively applied,
combined and bypassed on a per-client, per-recipient or per-sender basis,
allowing varying levels of severity and/or permissiveness.  Postfix
leverages simple lookup tables to support arbitrarily complicated
user-defined sequences of restrictions and exceptions, with policy
servers\footnote{Policy servers will be explained in \sectionref{policy
servers}.} as the ultimate in flexibility.  Administrators can also supply
their own rejection messages to make it clear to senders why exactly their
mail was rejected.  Unfortunately this flexibility has a cost: complexity
in the logs generated.  While it is easy to use standard Unix text
processing utilities to determine the fate of an individual email,
following the journey an email takes through Postfix can be quite
difficult.  The logs tend to follow a 90\%-10\% pattern: 90\% of the time
the journey is simple, but the other 10\% of the time requires 90\% of the
code.\footnote{These numbers don't have a solid scientific basis, they're
based on gut feeling from writing and debugging the software.}

Postfix's design follows the Unix philosophy of \textit{Write programs that
do one thing and do it well\/}~\cite{unix-philosophy}, and is separated
into various component programs to perform the tasks required of an
\MTA{}\@: receive mail, send mail, local delivery of mail, etc. --- full
details can be found at~\cite{postfix-overview}.  Each log line contains
the name of the Postfix program which produced it, and this information is
used when determining which rules should be used to parse each log line
(see \sectionref{rule characteristics} for details.)

@@

It should be clear from the earlier Postfix background (\sectionref{postfix
background}) that logs produced by Postfix are not fixed; they vary widely
from host to host, depending on the set of restrictions chosen by the
administrator.  With this in mind, one of the parser's design aims was to
make adding new rules as easy as possible, to enable administrators to
properly parse their log lines.  The main design decision which enables
this is the separation of rules from the parsing algorithm, and secondly
the division of the algorithm into discrete actions.

@@

Separating the rules from the algorithm makes it possible to parse new log
lines without modifying the core parsing algorithm.  Although this may seem
like a trivial point, is it substantially more difficult to understand a
program's entire parsing algorithm, identify the correct location to
change, and make the appropriate changes without adversely affecting
parsing, particularly as there may be edge cases which are not immediately
obvious.\footnote{See \sectionref{yet-more-aborted-delivery-attempts} for a
complication which occurs only four times in 93 log files tested.}
Requiring changes to the parsing algorithm also complicates upgrades, as
the changes must be preserved during the upgrade, and may clash with
changes made by the developer.  \parsername{} allows the user to add new
rules to the database without changing the parsing algorithm, unless the
new log lines to be parsed require functionality not already provided by
the algorithm.  If the new log lines do require new functionality, new
actions can be added to the parser without modifying existing actions or
other parts of the algorithm; only in the rare case that the new actions
require support from other sections of the code will more extensive changes
be required.

@@

Parsing in \LMA{} requires major changes to the code to parse new log lines
or extract additional data.  It does not appear to deal with any of the
complications discussed in \sectionref{complications} (initial
complications) and \sectionref{additional complications} (additional
complications).  It does not differentiate between different types of
rejections, so it is not suitable for the purposes of this project; the
data about which restriction caused the rejection is discarded, whereas an
important goal of this project is to retain that data to aid optimisation
and evaluation of Postfix restrictions.  \LMA{} fails to parse Postfix log
files generated on Solaris hosts because the fields added automatically
differ from those added on Linux hosts; logs from Solaris hosts (and
possibly other operating systems) thus require preprocessing before parsing
by \LMA{}.

@@

The database schema can be conceptually divided in two: the rules which are
used to parse log files, and the data saved from the parsing of log files.
Rules have the fields required to parse the log lines, extract data to be
saved, and the action to be executed; they also have several fields which
aid the user in understanding what each rule parses.  The rules are
described in detail in \sectionref{rules} but the fields are covered here.

The data saved from parsing the logs is also divided into two tables as
described below: connections and results.  The connections table contains a
row for every mail accepted and every connection where there was a
rejection; the individual fields will be described in the forthcoming
\sectionref{connections table}.  The results table has one or more rows per
row in the connections table, depending on the specifics of the
mail/connection; the fields will be covered in detail in
\sectionref{results table}.

@@

Rules are discussed in detain in \sectionref{rules}, but the rules table is
covered here.  Rules are created by the user, not the parser, and will not
be modified by the parser (except for the hits and hits\_total fields).
Rules parse the individual lines, telling the parser which fields to
extract and what action to take for that line.

@@

        For efficiency the keywords are expanded and every rule's \regex{}
        is compiled before attempting to parse the log file --- otherwise
        each \regex{} would be recompiled each time it was used, resulting
        in a large, data dependent slowdown.  Rule efficiency concerns are
        discussed in \sectionref{rule efficiency}.

@@

    \item [action] The action the algorithm will take; a full list can be
        found in \sectionref{actions-in-detail}.

@@

    \item [hits] is an efficiency measure.  This counter is maintained for
        every rule and incremented each time the rule successfully matches.
        At the start of each run the program sorts the rules in descending
        order of hits, and at the end of the run updates every rule's hits.
        Assuming that the distribution of log lines is reasonably
        consistent between log files, rules matching more commonly
        occurring log lines will be tried before rules matching less
        commonly occurring log lines, lowering the program's execution
        time.  Rule ordering for efficiency is discussed in
        \sectionref{rule ordering for efficiency}.

@@

    \item [priority] This is the user-configurable companion to hits: rules
        will be tried in order of priority, overriding hits.  This allows
        more specific rules to take precedence over more general rules
        (described in \sectionref{overlapping rules}).

@@

This section discusses the rules used in parsing Postfix log files,
starting with rule characteristics, continuing with detecting and dealing
with overlapping rules, and the problems they can cause.  An example rule
and a line it would match are provided, plus a description of how the
fields in the rule are used when matching and performing the action (the
details of the table containing the rules have already been described in
\sectionref{rule attributes},).  This section continues with a discussion
of rule efficiency concerns, referring to the graphs in
\sectionref{graphs}, and finishes with an explanation of the algorithm used
to generate a new \regex{} from unparsed lines.

Please refer to \sectionref{parser design} for a discussion of why the
rules and actions have been separated in the parser's design.

@@

    \item Compare the results of parsing using sorted, shuffled and
        reversed rules.\footnote{See \sectionref{rule efficiency} for more
        details of sorting the rules.}  Parse a number of log files using
        normal sorting, dump a textual representation of the connections
        and results tables, and delete everything from those tables.
        Repeat with shuffled and reversed sorting.  If there are no
        overlapping rules the tables from each run will be identical;
        differences indicate overlapping rules.  Which rules overlap can be
        determined by examining the differences in the tables: each result
        contains the id of the rule which created it, so the rule
        referenced in the normal table overlaps with the rule referenced in
        the reversed table.  Unfortunately this method cannot prove the
        absence of overlapping rules; it can detect overlapping rules, but
        only if there are log lines in the input files which match more
        than one rule.

@@

    \item [action:] will be executed if the \regex{} matches successfully
        (see \sectionref{actions-in-detail} for full details).

@@

    \item [hits, hits\_total and priority:] hits and priority are used in
        ordering the rules (see \sectionref{rule ordering for efficiency});
        hits is set to the number of successful matches at the end of the
        parsing run, and hits\_total has hits added to it, but is otherwise
        unused by the algorithm.

@@

Rule ordering was mentioned in \sectionref{rule attributes} and will be
covered in greater detail in this section.  At the time of writing there
are 139 different rules, with the top 10\% matching the vast majority of
the log lines, and the remaining log lines split across the other 90\% of
the rules (as shown in graph~\refwithpage{rule hits graph}).  Assuming that
the distribution of log lines is reasonably steady over time, program
efficiency should benefit from trying more frequently matching rules before
those which match less frequently.  To test this hypothesis three full test
runs were performed with different rule orderings:

@@

    \item [Pre-process the file]  The new algorithm leverages the knowledge
        gained while writing rules and performs a large number of
        substitutions on the input lines, replacing commonly occurring
        variable terms (e.g.\ email addresses, \IP{} addresses, the
        standard start of rejection messages, etc.) with \regex{} keywords
        which the parser will expand when it loads the rule (see the
        \regex{} entry in \sectionref{rule attributes}).  The purpose of
        this phase is to utilise the existing knowledge to create more
        accurate \regexes{}.  The new log lines are written to a temporary
        file, which all subsequent stages use instead of the original input
        file.

@@

    \item The database table containing the rules is dealt with in
        \sectionref{rule attributes}, and is not duplicated in this
        section.

@@

Please refer to \sectionref{parser design} for a discussion of why the
rules and actions have been separated in the parser's design.

@@

Multiple independent mails may be delivered during one connection: this
requires cloning the current data as soon as a mail is accepted, so that
subsequent mails won't trample over each other's data.  This must be done
every time a mail is accepted, as it's impossible to tell in advance which
connections will accept multiple mails.  It is quite easy to overlook this
complication because only a small minority of connections accept more than
one mail. Happily once the mail has been accepted log entries won't be
correlated by \pid{} for that mail any more (its queueid will be used
instead), so there isn't any ambiguity about which mail a given log line
belongs to.\footnote{Unfortunately this statement is not completely
accurate: see \sectionref{timeouts-during-data-phase} for details.  However
in general there isn't any ambiguity about which data structure should be
used for a given log line.}  The original connection will be discarded
unsaved when the client disconnects if it doesn't have any data worth
saving, i.e.\ no rejections.  One unsolved difficulty is distinguishing
between different groups of rejections, e.g.\ when dealing with the
following sequence:

@@

Figures~\refwithpage{flow chart image part 1} and~\refwithpage{flow chart
image part 2} shows the paths the data representing a mail/connection can
take through the parser algorithm.  The flow chart covers the most common
paths only; there are additional, uncommon paths which are excluded for the
sake of clarity, but are described in \sectionref{additional
complications}.

@@

Several more states and corresponding branches are omitted for the sake of
clarity (these extra states are caused by solving the difficulties
described in \sectionref{additional complications}).

The solution to the \textit{out of order log log lines\/} complication
encountered during development (\sectionref{out of order log lines})
requires that a list of the acceptable combinations of Postfix programs a
mail can pass through; however this does not correspond to the
\textit{actions\/} a mail must pass through.  Before that complication was
overcome, i.e.\ for the majority of the development of this parser, there
was nothing in the parser explicitly encoding the states or paths a mail
could take; the vast majority of the parser has no need for this flow of
data to be specified.  This eases the process of adding new actions
(described in \sectionref{adding new actions}), as there is no requirement
to explicitly insert the action into an algorithmic version of the flow
chart below.

@@

\sectionref{actions-in-detail} explains the actions in substantive detail;
this section omits such detail because it would clutter and confuse the
algorithm description.  The flow chart in \sectionref{flow-chart} should
also be consulted while reading this section.  For the sake of clarity this
description covers only the most common paths through the algorithm, as
including every path would hinder understanding; the more uncommon paths
are caused by the complications described in \sectionref{additional
complications}, and are covered therein.

@@

    \item If one or more mails were accepted there will be more log entries
        for those mails later, see \sectionref{mail-delivery}.

@@

Re-injection due to forwarding sadly lacks explicit log lines of its
own;\footnote{Previously discussed in \sectionref{complications},
complication 3.} re-injection is somewhat awkward to explain because it
overlaps both the mail acceptance and mail delivery sections, so discussion
is deferred to \sectionref{tracking re-injected mail}.

Internally generated mails lack any explicit origin in Postfix 2.2.x and
must be detected using heuristics (see
\sectionref{identifying-bounce-notifications} for details).  Bounce
notifications are the primary example of internally generated mails, though
there may be other types.

@@

    \item Sometimes mail is re-injected and the child mail needs to be
        tracked by the parent mail (action: track; flowchart:~10) ---
        \sectionref{tracking re-injected mail} discusses this in
        detail.

    \item Eventually the mail will be delivered, bounced, or deleted by the
        administrator (action: commit; flowchart:~12).  This is the last
        log line for this particular mail (though it may be indirectly
        referred to if it was re-injected).  If it is neither parent nor
        child of re-injection the data is cleaned up and entered in the
        database (flowchart:~14), then deleted from the state tables.
        Re-injected mails are described in \sectionref{tracking re-injected
        mail}.

@@

It should be reiterated that the actions above happen whether the mail is
delivered to a mailbox, piped to a command, delivered to a remote server,
bounced (due to a mail loop, delivery failure, or five day timeout), or
deleted by the administrator.  The exception is what happens after delivery
to the parent or children of mail re-injected due to forwarding, as
explained in \sectionref{tracking re-injected mail}.

@@

    \item The indicator which more commonly introduces re-injection is when
        \daemon{qmgr} selects a mail with a previously unseen queueid for
        delivery (action: mail\_picked\_for\_delivery; flowchart:~3), in
        which case a new data structure will be created.  The mail will be
        flagged as having unknown origins; this flag should be subsequently
        cleared once the origin has been established.  This may also be an
        indicator that the mail is a bounce notification, see
        \sectionref{identifying-bounce-notifications} for details.

@@

    \item [EXPIRY] If Postfix has not managed to deliver a mail after
        trying for five days it will give up and return the mail to the
        sender.  When this happens the mail will not have a combination of
        Postfix programs which passes the valid combinations check (see
        \sectionref{out of order log lines}).  To ensure that the mail can
        be committed the EXPIRY action sets a flag marking the mail as
        expired; the flag later causes the valid combinations check to be
        skipped, so the mail will be committed.

@@

    \item [MAIL\_PICKED\_FOR\_DELIVERY] This action represents Postfix
        picking a mail from the queue to deliver. This action is used for
        both \daemon{qmgr} and \daemon{cleanup} as it needs to deal with
        out of order log lines; see \sectionref{discarding cleanup lines}
        for details.

@@

    \item [PICKUP] The PICKUP action corresponds to the \daemon{pickup}
        service dealing with a locally submitted mail.  Out of order log
        entries may have caused the state table entry to already exist (see
        \sectionref{pickup logging after cleanup}); otherwise it is
        created.  The data from the log line is then saved to the state
        table entry.

@@

    \item [TIMEOUT] The connection timed out so the mail currently being
        transferred must be discarded. The mail may have been accepted, in
        which case there's a data structure to dispose of, or it may not in
        which case there is not.  See
        \sectionref{timeouts-during-data-phase} for the gory details.

@@

Adding new actions is almost as simple as adding new rules, though actions
are not stored in the database.  The implementor writes a subroutine which
accepts the standard arguments given to actions, and registers it as an
action by calling the parser subroutine add\_actions() with the name of the
new action subroutine as a parameter.  No other work is required on the
implementor's part; all of their attention and effort can be focused on the
correctness of their action.  The only negative aspect is that the process
involves editing the parser source code, which makes upgrading to a later
version of the parser more difficult, though by no means impossible.  If
the author of the new action wishes they can take advantage of the parser's
object oriented implementation~\cite{Wikipedia-object-orientation} by
subclassing it and implementing their changes in the derived class,
allowing future upgrading of the parser with greatly reduced chance of
conflicts.\footnote{The real difference between the two approaches is where
the new code is placed.  The simpler option is to change the parser code
directly, but those changes will then have to be made to subsequent
versions of the parser, and as the scope of the changes increases so does
the chance of conflict.  The more time consuming option is to write a
subclass containing the new actions and change the program which invokes
the parser so that it uses the subclass rather than the parser; the changes
required to the program invoking the parser are minor and much less likely
to lead to conflicts when upgrading to future versions of the parser.  An
alternative is to submit new actions to the author of the parser for
inclusion in future versions, resulting in two benefits: the new actions do
not need to be maintained separately, and other users of the parser will
avail of the new functionality.} The action may need to extend the list of
valid combinations described in \sectionref{out of order log lines} if the
addition creates a different set of acceptable programs, but this is
unlikely to occur, as it would require parsing log lines from Postfix
components the parser currently ignores.\footnote{The mail server used for
development does not utilise either the \daemon{lmtp} or \daemon{virtual}
delivery agents, so this parser does not have rules to handle log lines
from those components.  Adding new rules to parse those component's log
lines is a simple process, though if their behaviour differs significantly
from the \daemon{smtp} or \daemon{local} delivery agents new actions may be
required.  The mail server in question is a production mail server handling
mail for a university department; the benefit is that the logs used exhibit
the idiosyncrasies and foibles a mail server in the wild must deal with,
but the downside is that significantly altering the configuration just to
log messages from a different Postfix component is not an option.}

@@

Postfix 2.2.x (and presumably previous versions) lacks explicit logging
when bounce notifications are generated; suddenly there will be log entries
for a mail which lacks an obvious source.  There are similarities to the
problem of re-injected mails discussed in \sectionref{tracking re-injected
mail}, but unlike the solution described therein bounce notifications do
not eventually have a log line which identifies their source.  Heuristics
must be used to identify bounce notifications, and those heuristics are:

@@

This complication is further complicated by the presence of out of order
\daemon{cleanup} lines: see \sectionref{discarding cleanup lines} for
details.

@@

    \item The appropriate action to take --- discussed in
        \sectionref{mails-covered}.
